---
title: "Capstone Slides"
author: "Pingping Zhou"
date: "2025-11-16"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## R Markdown

##This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

##When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Output

```{r}
install.packages(c(
  "mgcv", "pROC", "caret", "dplyr", "ggplot2", "scales",
  "rmarkdown", "knitr", "kableExtra", "xaringan", "xaringanthemer"
))
```
# Save this as: slides.Rmd
# Then run: rmarkdown::render("slides.Rmd")

---
title: "GAM Fraud Detector: Interpretable Risk Modeling"
author: "Your Name"
date: "November 16, 2025"
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts]
    nature:
      highlightStyle: github
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, 
                      fig.width = 8, fig.height = 5, dpi = 300)
library(dplyr); library(ggplot2); library(mgcv); library(pROC); library(caret)
library(scales); library(kableExtra)

**Primary Goal**
To identify fraudulent transactions by analyzing patterns in transaction characteristics and assigning fraud probabilities. The model aims to maximize fraud detection (sensitivity) while minimizing false alarms (false positives), which is critical in financial fraud prevention.




## Slide with Plot

##```{r}

##```


