---
title: "Generalized Additive Models in Fraud Detection"
subtitle: "Data Science Capstone Project"
author: "Grace Allen, Kesi Allen, Sonya Melton, Pingping Zhou"
date: "November 21, 2025"
format: 
  html:
    code-fold: true       # Allows folding of code chunks
    self-contained: true  # Packages images and CSS in the HTML
    toc: true 
output-dir: report1      # Folder where rendered HTML will go
bibliography: references.bib
csl: apa.csl 
execute: 
  warning: false
  message: false
css: styles.css 
course: "Capstone Projects in Data Science"
editor:
  markdown: 
    wrap: 72
---


::: callout-important
**Remember:** Your goal is to make your audience understand and care
about your findings. By crafting a compelling story, you can effectively
communicate the value of your data science project.

Carefully read this template since it has instructions and tips to
writing!

More information about `revealjs`:
<https://quarto.org/docs/reference/formats/presentations/revealjs.html>
:::

## Introduction  {.smaller}

-   Develop a storyline that captures attention and maintains interest.

-   Your audience is your peers

-   Clearly state the problem or question you're addressing.

-   Introduce why it is relevant needs.

-   Provide an overview of your approach.

In kernel estimator, weight function is known as kernel function
[@efr2008]. Cite this paper [@bro2014principal]. The GEE [@wang2014].
The PCA [@daffertshofer2004pca]\*

## Methods

-   Detail the models or algorithms used.

-   Justify your choices based on the problem and data.



## Data Exploration and Visualization

-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.
##Data set Description
🔹 What It Is
•	A synthetic dataset built to mimic real financial transactions
•	Privacy safe: No actual personal info is included, so it's fully private and ethical.
•	Hosted on Kaggle
🔹Why We Use It
•	Train fraud detection models for binary classification tasks
•	Spot fraud: each transaction labeled as fraud (1) or not fraud (0)
🔹 What Makes It Special:
•	Realistic Fraud: It includes tricky patterns like:
o	Groups of fraudulent transactions.
o	Small, hard-to-notice unusual activities.
o	Odd user behaviors.
•	Big & Diverse: Large enough with a mix of normal and rare fraud cases to help with imbalanced data challenges.

##Key Characteristics
🔹 What's Inside & What We Do
•	50,000 Rows: A good amount of data to work with.
•	Two Labels: Every transaction is marked as either:
o	1 = Fraud
o	0 = Not Fraud
🔹Data Features-- 21 features across three categories:
•	Numbers: Like transaction amounts, risk scores, account balances.
•	Categories: Transaction types (payment, transfer, withdrawal), device types, merchant categories.
•	Time Data: When transactions happened (time, day) and their sequence.
🔹Label Distribution & Realism
•	Class Imbalance: Fraudulent transactions are a small percentage, reflecting real-world scenarios.
•	Behavioral Realism: Includes unusual spending, behavioral signals, and high-risk profiles.
•	Modeling flexibility: supports interpretable (GAMs, logistic regression) or high-performance (XGBoost) approaches


## Data Exploration and Visualization {.smaller}
The charts (including Tables 1, 2, and 3) show that all categories in the dataset — transaction types, device types, and merchant groups — are almost perfectly balanced. Each transaction type has about 12,500 cases, each device type around 16,600–16,800, and each merchant category close to 10,000. This even distribution makes it simple to compare counts and spot patterns. Generalized Additive Models (GAMs) are still a strong choice because they can handle both numbers and categories, and capture complex, nonlinear relationships in the data.

```{r, warning=FALSE, echo=F, message=FALSE}
# Load libraries
library(tidyverse)
library(janitor)
library(gt)
library(scales)

# === Load dataset ===
data_path <- "synthetic_fraud_dataset.csv"
df <- readr::read_csv(data_path, show_col_types = FALSE) |>
  clean_names()

# === Create count tables ===
tbl_type <- df |>
  count(transaction_type, name = "Count") |>
  arrange(desc(Count)) |>
  rename(Type = transaction_type)

tbl_device <- df |>
  count(device_type, name = "Count") |>
  arrange(desc(Count)) |>
  rename(Device = device_type)

tbl_merchant <- df |>
  count(merchant_category, name = "Count") |>
  arrange(desc(Count)) |>
  rename(Merchant_Category = merchant_category)

# === Blue Theme for gt Tables ===
style_blue_gt <- function(.data, title_text) {
  .data |>
    gt() |>
    tab_header(title = md(title_text)) |>
    fmt_number(columns = "Count", decimals = 0, sep_mark = ",") |>
    tab_options(
      table.font.names = "Arial",
      table.font.size  = 14,
      data_row.padding = px(6),
      heading.align    = "left",
      table.border.top.color    = "darkblue",
      table.border.top.width    = px(3),
      table.border.bottom.color = "darkblue",
      table.border.bottom.width = px(3)
    ) |>
    tab_style(
      style = list(cell_fill(color = "darkblue"),
                   cell_text(color = "white", weight = "bold")),
      locations = cells_title(groups = "title")
    ) |>
    tab_style(
      style = list(cell_fill(color = "steelblue"),
                   cell_text(color = "white", weight = "bold")),
      locations = cells_column_labels(everything())
    ) |>
    opt_row_striping() |>
    cols_align("right", columns = "Count")
}

# === Render all three blue tables ===
style_blue_gt(tbl_type, "Table 1 – Transaction Types and Counts")
style_blue_gt(tbl_device, "Table 2 – Device Types and Counts")
style_blue_gt(tbl_merchant, "Table 3 – Merchant Categories and Counts")
```

Figure 1. Histograms show key variables used in fraud detection. Most features—Account_Balance, Transaction_Distance, Risk_Score, and Card_Age—appear uniformly distributed, supporting nonlinear modeling such as GAM. Transaction_Amount is strongly right-skewed, with many small transactions and a few large outliers, indicating that extreme amounts may signal potential fraud. This skewness suggests that transformations (e.g., log-scale) or nonlinear models like GAM can better capture fraud patterns and improve prediction accuracy.

**Distribution of Variables**

```{r}
library(tidyverse)
library(lubridate)
library(patchwork)  # for arranging multiple ggplots

# Load dataset
fraud_data <- read.csv("synthetic_fraud_dataset.csv")

# Convert Timestamp to date and calculate Issuance_Year if needed
fraud_data <- fraud_data %>%
  mutate(
    Timestamp = ymd_hms(Timestamp, quiet = TRUE),  # adjust format if needed
    Transaction_Year = year(Timestamp),
    Issuance_Year = Transaction_Year - Card_Age
  ) %>%
  filter(!is.na(Card_Age))  # remove rows with NA in Card_Age

# Variables to plot (move Transaction_Amount to last)
numeric_vars <- c("Account_Balance", "Transaction_Distance", "Risk_Score", "Card_Age", "Transaction_Amount")

# Create a list to store plots
plot_list <- list()

# Generate plots and store in the list
for (var in numeric_vars) {
  p <- ggplot(fraud_data, aes_string(x = var)) +
    geom_histogram(fill = "steelblue", color = "white", bins = 30) +
    labs(title = paste("Distribution of", var),
         x = var,
         y = "Count") +
    theme_light()
  
  plot_list[[var]] <- p
}

# Arrange plots in a grid: 2 plots per row
(plot_list[[1]] | plot_list[[2]]) /
(plot_list[[3]] | plot_list[[4]]) /
plot_list[[5]]  # Transaction_Amount appears last

```


Figure 2 shows that most cards are new, and older cards are less common. Older cards may be more at risk because their security features may be outdated, while newer cards may have different usage patterns, like more digital transactions. These differences suggest that fraud risk changes with card age, and models like GAM can help capture those patterns.
```{r}
library(tidyverse)
library(lubridate)
# Load libraries
library(ggplot2)
library(dplyr)
library(tidyr)    # For pivot_longer
library(gridExtra) # For arranging plots
#install.packages("moments") 
library(moments)   # For skewness and kurtosis
# Load dataset
fraud_data <- read.csv("synthetic_fraud_dataset.csv")

# Convert Timestamp to date, calculate Transaction Year and Issuance Year, exclude NAs
fraud_data <- fraud_data %>%
  mutate(
    Timestamp = ymd_hms(Timestamp),               # adjust if format differs
    Transaction_Year = year(Timestamp),
    Issuance_Year = Transaction_Year - Card_Age
  ) %>%
  filter(!is.na(Issuance_Year), !is.na(Card_Age))  # remove rows with NA

# Bin Issuance Year into 5-year ranges and drop unused NA factor levels
fraud_data <- fraud_data %>%
  mutate(
    Issuance_Year_Bin = cut(Issuance_Year,
                             breaks = seq(2000, 2025, by = 5),
                             right = FALSE,
                             labels = c("2000-2004","2005-2009","2010-2014","2015-2019","2020-2024"))
  ) %>%
  filter(!is.na(Issuance_Year_Bin))  # drop any rows that fall outside the bins

# Histogram
ggplot(fraud_data, aes(x = Issuance_Year_Bin)) +
  geom_bar(fill = "steelblue", color = "white") +
  labs(title = "Figure 2. Card Age by Issuance Year Range",
       x = "Card Issuance Year Range",
       y = "Count") +
  theme_light()

```

Figure 3 shows that fraud probability changes in a nonlinear way with transaction amount. Small payments usually have low risk, while very large amounts show higher risk. This makes transaction amount a key predictor and shows why GAMs are useful — they can flexibly capture these curved patterns.

```{r}
library(tidyverse)
# Load dataset
fraud_data <- read.csv("synthetic_fraud_dataset.csv")
# Ensure Fraud_Label is numeric (0/1)
fraud_data <- fraud_data %>%
  mutate(Fraud_Label = as.numeric(Fraud_Label))

# Nonlinearity check: Transaction Amount vs Fraud Probability
ggplot(fraud_data, aes(x = Transaction_Amount, y = Fraud_Label)) +
  geom_smooth(method = "loess", se = FALSE, color = "darkblue") +
  labs(title = "Figure 3. Transaction Amount and Fraud Probability",
       x = "Transaction Amount",
       y = "Fraud Probability") +
  theme_light()
```




## Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

## Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
